---
title: "Contrasts, R, and regression"
author: "Serje Robidoux"
date: "04/08/2021"
output: pdf_document
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(echo = TRUE)


library(dplyr); library(emmeans)
```

## Contrasts using contr.* vs emmeans
Using contrasts in models like *lm/glm*, *lmer/glmer*, etc is not as straightforward as it seems.

### The short version
The main problem arises because while you can specify the contrasts on data directly using the *contr.\*()* family of contrast functions, *lm* etc... actually use the inverse of the resulting matrix. So when you specify treatment contrasts using *contr.treatment(4)*, *lm* actually fits something quite different.

## Demonstrating the relationships
Let's start with demonstrating the problem on some artificial data. I will create data with 4 cells, each with means of 100, 200, 300, and 400, with SDs of 20. So the Grand Mean is 250.

```{r fakeDat}
set.seed(1) # Just makes sure the randomization stays the same on different runs
# Create some fake data with four conditions
fakedat = expand.grid(subj=1:10, cond=c("a", "b", "c", "d"), Items=1:10) %>% 
  group_by(cond) %>% 
  mutate(dv = scale(rnorm(n()))*20 + 100*as.numeric(cond)) %>% ungroup()
# this produces means for each condition:

kable(merge(aggregate(dv~cond, fakedat, mean), aggregate(dv~cond, fakedat, sd)
      , by="cond", suffixes=c(".m", ".sd")))

```


## R's default behaviour

#### lm vs glm
I'm going to stick with lm because it doesn't transform the dependent variable the way glm does. This allows us to easily and directly compare the coefficients we get out of the regression to the means for each condition. With glm we'd have to do a lot more work to match it up with the mean accuracies, etc...

Now let's look at what happens if we do nothing but fit a simple linear regression model.:

```{r defaultRegression}

summary(lm(dv~cond, fakedat))
```

You can see that the intercept is 100 (the mean of condition 'a'), and the estimate for each cond ('b', 'c', 'd') is the difference between that condition's mean and 100 (a's mean).

By default, R uses something called "treatment" contrasts. You can see this by looking at the contrasts on 'cond': 

```{r}
kable(contrasts(fakedat$cond))
```

and comparing it to 
```{r}
kable(contr.treatment(4))
```

So note that even though this doesn't look like $B-A$, $C-A$, and $D-A$, that is what the resulting output represents. The reason is that the math of regression means that the **estimates** are actually generated by the *inverse* of the contrast matrix.

```{r invTreatmentContrasts}

# there's a lot going on in here, including adding an "intercept" contrast.

kable(t(solve(cbind(rep(1,4), contr.treatment(4)))))

```

the rep(1,4) thing is me adding a contrast to represent the intercept (1,1,1,1). As a result there are 4 "contrasts" in the inverted matrix - the first is the intercept, and the next ones are the ones related to the condition variable.

Looking at that matrix, you can see that the first estimate (the intercept) is just $A$, and the next three estimates (2,3,4) are $B-A$, $C-A$, $D-A$.

## contr.sum

When you call *contr.sum* it looks like the matrix you want:

```{r}
kable(contr.sum(4))
```

Which is very similar to the result of taking the inverse of *contr.treatment()*. The only difference is that the last category (D) is now the "reference condition" BUT, remember that if you specify a contrast, functions like lm/glm etc... are going to estimate the "inverse".

And the inverse of the *contr.sum()* matrix is a very different beast:

```{r invSumContrasts}
# there's a lot going on in here, including adding an "intercept" contrast.

kable(t(solve(cbind(rep(1,4), contr.sum(4)))))

```

(The first column is just a grand mean $(A+B+C+D)/4$, which means that the intercept is going to represent the grand mean)

### What is contr.sum?

Recall in our data that the grand mean was 250, and the means were 100, 200, 300, 400. Here's what happens if we tell R to use sum contrasts when fitting the data:

```{r sumContrastRegression}

contrasts(fakedat$cond)=contr.sum(4)
summary(lm(dv~cond, fakedat))
```

The Intercept is the grand mean, as expected, and the next three values are: "$100-250=-150$ is $A-GrandMean$", "$200-250 = -50$: $B - GrandMean$", "$300-250 = 50$: $C-GrandMean$"

[Note: You do not have a test of D, because you can only do three tests.]

It's not readily apparent that the resulting matrix is comparisons to the grand mean, but it can be clearer if we consider $A - GrandMean$. If we wanted to calculate that we would use the following:

$A - (A+B+C+D)/4 = A - .25A -.25B-.25C-.25D = .75A -.25B-.25C-.25D$

So to get that difference, you use the coefficients $(.75, -.25, -.25, -.25)$, which is what the inverse of *contr.sum(4)* is producing.

## Q: What if I want $A - (A+B)/2$?

As Lyndsey pointed out, if you take the formula above and look at just two conditions:

$A - (A+B)/2 = .5A-.5B$, so we want coefficients $(-.5, .5)$ in the estimate.

To get at the right contrast matrix, though, we have to find its inverse.

So the contrast matrix we would have to specify is:

```{r invSumEstimates}

kable(t(solve(cbind(c(1,1), c(.5, -.5)))))

```

Which is the *contr.sum(2)* (column 2; column 1 is the intercept)

which leads to...

## Q: Why do people say you can use either (1, -1) or (.5, -.5)?

Because the only difference will be scaling. The estimate will be twice as large, but you will get the same t-statistic and p-value. Since we are usually not interested in the exact size of the estimate and only its significance, it doesn't make any difference to our conclusions.

```{r comparingScalings}
smalldat = fakedat %>% filter(cond %in% c('a', 'b'))
smalldat$cond = factor(smalldat$cond) # get rid of "unused" levels c,d
kable((contrasts(smalldat$cond) = contr.sum(2)))
summary(lm(dv~cond, smalldat))

# compare that to cutting the contrasts in half:
kable((contrasts(smalldat$cond) = contr.sum(2)/2))
summary(lm(dv~cond, smalldat))

```

The only difference is that using .5, -.5 has doubled the difference estimate and the SE. The t-stat and p-value stay the same.

It does, however, make a difference if you want to make the estimates match up with your means. We expect the difference to be 100, so only the ".5, -.5" gets it right.  The reason is that the inverse of *contr.sum(2)/2* (.5, -.5) is the same as the inverse of *contr.treatment(2)* (at least for that comparison):

```{r twoConditionCase}

kable(t(solve(cbind(c(1,1), contr.sum(2)/2))))
kable(t(solve(cbind(c(1,1), contr.treatment(2)))))
```

The intercepts are different, but the estimate for "cond" is the same.

The fact that you can use a "scaled sum contrast" to get the same result as a "treatment contrast" is just a quirk of the 2 conditions scenario. It does not generalize, so it's probably best to just forget it and use *contr.treatment(2)*. (Note, I have told people to just specify .5, -.5 in the past, but now I prefer to use *cont.\** or *emmeans*)

## Oksana's matrices

In Oksana's last email, she described three matrices:

```{r OLMatrices}

kable((mat1 = contr.sum(4)))
kable((mat2 = contr.treatment(4)))
kable((mat3 = matrix(c(   .75, -.25, -.25
                       , -.25, -.25,  .75
                       , -.25, -.25, -.25
                       , -.25,  .75, -.25), nrow=4)))
```

The first two we've already covered above. If we use the 1,0,0,-1 construction (*contr.sum()*) we end up with the "$A - GrandMean$" type of output. If we use 1,0,0,0 (*contr.treatment()*), we get the $B-A$, $C-A$, $D-A$ comparisons.

What happens if you use the third matrix? It turns out you get something similar to *contr.treatment*: $A-D=-300$, $B-D=-200$, $C-D=-100$ output, but now the intercept is the GrandMean (250), rather than just $D$ and the reference condition is $D$ instead of $A$:

```{r OLmat3}
contrasts(fakedat$cond)=mat3
summary(lm(dv~cond, fakedat))
```

###Why does this happen?

Well, if we invert that matrix, what we get is the following:

```{r invMat3}
kable(t(solve(cbind(rep(1,4), mat3))))
```

Ignore the first component for a moment: the three contrasts are nearly identical to when we inverted *contr.treatment()* other than now using $D$ as the reference level instead of A - so for those three contrasts we get $D-A$, $D-B$, $D-C$.

The first component (as always, the intercept) is now the average of all 4 levels, much as it was with *contr.sum()*. So that matrix ends up being a sort of hybrid between sum and treatment contrasts (or what Sundström refers to as dummy and effect coding).


#### Why do I keep adding an intercept [c(1,1,1,1)]?

Two reasons: 1) *lm* also has to have a contrast for the intercept, but you don't specify it because it is *always* a list of 1s (ie, 1,1,1,1), and 2) we can't invert a matrix unless it is square (eg, 4x4).

But the interpretation of one contrast is not independent of the other contrasts, so it's particularly important to use the intercept or it wouldn't make sense.

Eg,

```{r squareMat}
kable(cbind(c(1,1,1,0), contr.treatment(4)), caption="contrast")
kable(t(solve(cbind(c(1,1,1,0), contr.treatment(4)))), caption="inverse")
```

With that weird version, we seem to be getting $A$, $B-A$, $C-A$, and $D$