---
title: "Mixture Modelling and Latent Profile Analysis"
author: "Serje Robidoux"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
include(tidyverse); include(gridExtra);include(mvtnorm);include(plot3D)
knitr::opts_chunk$set(echo = FALSE)
set.seed(1850863759)
fig=0
tbl=0
grpCols = c("#FF008888", "#08FF8088", "#0055FF55", "#AA00FF55", "#FF880088")
```

# Mixture Modelling

The basic idea of mixture modelling is that when you are looking at distributions of data, you don't always know if you are looking at data from one population, or from many.

Mixture modelling attempts to uncover if there are actually several distinct populations in your data.

An example:

## Single variable

Consider that you have a data set (N=1000) that, when you plot the histogram, it looks like this:

```{r univariateMix, fig.cap="Histogram of a single variable with an estimated normal distribution"}
dat = rnorm(1000, mean=c(0, 1), sd = c(1, .2))

plot(hist(dat, breaks=50, plot=F), freq=F, main="Histogram", xlab="")
lines(x=seq(-4,4, length=1000), y=dnorm(seq(-4,4, length=1000), mean(dat), sd(dat)))
```
What might be going on here? The mean of the data is `r round(mean(dat),2)` with a standard deviation of `r round(sd(dat),2)`, but the "implied normal distribution" doesn't describe the data well. There appears to be a peak around 1, but a lot of data that is very low.

One possibility is that the data are taken from two different groups, that have different means and variances.  In this case, that is exactly right. I randomly sampled 500 data points from a Normal distribution with a mean of 0 and standard deviation of 1; and another 500 from a Normal distribution with a mean of 1 and a standard devation of .2. Figure `r (fig=fig+1)` depicts these two distributions, and what the distribution looks like when you sample from the two populations *without knowing that there are two distributions*.

```{r twoDists, fig.cap="The populations, and their combined distribution", fig.height=2}
dists = expand.grid(dist = 1:2, x = seq(-3,3,length=1000)) %>% 
  mutate(mu = dist-1
         , sigma = if_else(dist==1, 1, .2)
         , d = dnorm(x, mu, sigma)
         , dist = factor(dist))
univ = aggregate(d~x, dists %>% select(x,d), mean)


grid.arrange(ggplot(dists, aes(x=x, y=d, col=dist))+
  geom_line()+
  SOS$Theme+SOS$Colour+theme(legend.position="none")

, ggplot(univ, aes(x=x, y=d))+
  geom_line(col=sos.cols["purple"])+
  SOS$Theme+SOS$Colour
, nrow=1)

```

Figure `r (fig=fig+1)` compares that combined distribution to our histogram.

```{r mixModelDist, fig.cap="The mixed model distribution and the data."}

dists = expand.grid(dist = 1:2, x = seq(-4,4,length=1000)) %>% 
  mutate(mu = dist-1
         , sigma = if_else(dist==1, 1, .2)
         , d = dnorm(x, mu, sigma)
         , dist = factor(dist))
univ = aggregate(d~x, dists %>% select(x,d), mean)


hist(dat, main="Histogram", xlab="", breaks=50, freq=F)
lines(x=seq(-4,4, length=1000), y=univ$d, col=sos.cols["purple"]
      , lwd=2)

```


Unsurprisingly, that distribution does a much better job of capturing the shape of the data.

This was a toy example, though. Usually, you don't know:

1. Whether or not there is more than one underlying distribution generating your data;
2. how many underlying distributions there are; or
3. what each of those distributions look like.

Also, in most cases you are not using only a single variable. You might have dozens of variables creating large "multivariate" distributions.

Mixture Modelling is a set of methods that are used to try and "uncover" the answer to those three questions.

\newpage
## A bivariate example

Now let's look at what happens when you have two variables that are correlated around .8.

Figure `r (fig=fig+1)` depict the data if this was from a single population. You can see from the 3D plot that it sort of looks like a mountain. And the 2D plot is the scatterplot but with colours to indicate how much data is in each section.


```{r bivariateSample, fig.cap = "3D Histogram and 2D plot of some bivariate data."}

dat = rmvnorm(100000
              , mean=c(0, 1, 1.5)
              , sigma=matrix(
                c(            1,          .7*sqrt(.2), .5*sqrt(.5)
                  , .7*sqrt(.2),                   .2, .6*sqrt(.5)*sqrt(.2)
                  , .5*sqrt(.5), .6*sqrt(.5)*sqrt(.2), .5)
                , nrow=3)) %>% data.frame %>% 
  mutate(x = round(X1, 1), y=round(X2, 1), z=round(X3, 1))

grps_z = table(dat$x, dat$y)

par(mfrow=c(1,2))
hist3D(z=grps_z, phi=45, theta=10, x = as.numeric(rownames(grps_z))
       , y=as.numeric(colnames(grps_z)))
image2D(grps_z, x = as.numeric(rownames(grps_z)), y=as.numeric(colnames(grps_z)))

```

So an ideal bivariate normal distribution from a single population looks like a nice smooth mountain or hill.
\newpage

## A bivariate example, but with a mixture of samples.

Instead of a single group, this next figure is generated from 3 groups of people who come from different profiles of reading and math scores. (Note this is all randomly generated toy data.)

```{r multiMix, fig.cap = "3D histogram of a mixture of different bivariate data."}

# Randomly select the # of distributions:
ndists = 3
smps = sample(250:10000, ndists)
centres = matrix(runif(ndists*3,-3,3)^2, ncol=3)# %>% data.frame
vars = matrix(runif(ndists*3,.5,1.5), ncol=3) %>% data.frame %>% 
  mutate(covX1X2 = runif(ndists, min=-.5*sqrt(X1*X2), max=.8*sqrt(X1*X2))
         , covX1X3 = runif(ndists, min=-.5*sqrt(X1*X3), max=.8*sqrt(X1*X3))
         , covX2X3 = runif(ndists, min=-.5*sqrt(X2*X3), max=.8*sqrt(X2*X3)))

grps = vector("list", ndists)

for(i in 1:ndists)
  grps[[i]]=rmvnorm(smps[i], centres[i,1:3], 
                    matrix(unlist(vars[i,c(1,4,5,4,2,6,5,6,3)]), nrow=3)) %>% 
  data.frame(Group=i)

grps = grps %>% bind_rows %>% mutate(x=round(X1, 1), y=round(X2,1), z=round(X3,1))

grps_z = table(grps$y, grps$z)

par(mfrow=c(1,2))
hist3D(z=grps_z, phi=45, theta=180, x = as.numeric(rownames(grps_z))
       , y=as.numeric(colnames(grps_z)))
image2D(grps_z, x = as.numeric(rownames(grps_z)), y=as.numeric(colnames(grps_z)))

```

Now instead of a single smooth mountain, the 3D plot in Figure `r (fig=fig+1)` kind of looks like a mountain range, with mountains of different heights and shapes

# Latent Profile Analysis

LPA is just mixture modelling. The idea is that we are looking for ways to split the data up into different sets of "bivariate normal distributions" that can effectively reproduce the distributions in our observed data.

The way most LPA analyses work is to just "guess" a bunch of times and try to figure out which guesses do the best job of matching your data. They guess in very principled ways, and penalize overly complex models, but it is basically just guessing.

So let's have a look at an example of that process on our simulated 3 group data.

## LPA with mclust

I'm going to use the mclust package for this because it just seems the most intuitive to me.

```{r mclust, echo=T, message=F, fig.cap="Results from Mclust"}
library(mclust)
colnames(grps) = c("NeedForCog", "Reading", "Math", "Group", "z", "x", "y")
grps.LPA = grps %>% select(Reading, Math)
# our data is in a dataset called grps, and we only want to use Reading and Math
# There is a third variable in grps called NeedForCog that we'll discuss later.

# the first step is to just let mclust do all of the guessing.
# mclustBIC() returns how well the best guess did for:
# A: different numbers of groups (how many mountains?)
# B: different assumptions about the features of the groups (shapes of the mountains)
# This is slow, so I'll store the object to avoid running it over and over again.

if (file.exists("grps.BIC.RDat")) {
  load("grps.BIC.RDat")
} else {
  grps.BIC = mclustBIC(grps.LPA %>% scale)
  save(grps.BIC, file="grps.BIC.RDat")  
}
plot(grps.BIC)

```

mclustBIC() does a bunch of guessing based on how many groups (components) there might be, and each line/shape/colour represents different assumptions about what the shapes of those groups might be.

 Figure `r (fig=fig+1)` represents the model fits for different numbers of groups and kinds of models (more on this later). Higher values indicate better fits. There are clearly some models that do better than others and it seems like we need at least 3 groups. But it's not obvious that any one model is dominant. We can look at the actual results and get a bit more guidance.

```{r mclustBIC, echo=T}
grps.BIC
```

What we get here are the measures of how well the "best" fit was for each kind of model, and for 1-9 different groups. The larger the BIC value (careful with negative values), the better the fit. We also get a list of the best fitting models - the analysis seems to favour some pretty complex models (VVV with 3 to 5 groups).

We'll return to the letters later, so leave them aside for the moment. Let's just look at how each model is splitting up the data (Figure `r (fig=fig+1)`).

```{r mclustBest, echo=T, fig.cap = "Depictions of how the three best-fitting models split the data up", fig.height = 5}
best3 = lapply(strsplit(attr(summary(grps.BIC), "names"), ",")
               , function(x) rbind(x))

rng = c(-4,11)
par(mfrow=c(1,3))
grps.mclst = Mclust(grps.LPA, model=best3[[1]][1], G=best3[[1]][2], x = grps.BIC)
plot(grps.mclst, what="classification", ylim=rng, xlim=rng, col=grpCols)
title(main=paste0(best3[[1]][1],", ",best3[[1]][2]))

grps.mclst = Mclust(grps.LPA, model=best3[[2]][1], G=best3[[2]][2], x = grps.BIC)
plot(grps.mclst, what="classification", ylim=rng, xlim=rng, col=grpCols)
title(main=paste0(best3[[2]][1],", ",best3[[2]][2]))

grps.mclst = Mclust(grps.LPA, model=best3[[3]][1], G=best3[[3]][2], x = grps.BIC)
plot(grps.mclst, what="classification", ylim=rng, xlim=rng, col=grpCols)
title(main=paste0(best3[[3]][1],", ",best3[[3]][2]))


```


We can see right away that the three "best fitting" models are fairly similar, mostly differing in how many groups there are, but the general structure remains the same. 

The black circles with lines are there to describe the position, size, and direction of each of the distributions. One thing to note in the 5 group version is that there is a lot of overlap in the "higher math" distributions. So much so that there is a freen group that is almost entirely obscured by the purple and blue groups.

Next we'll discuss what those letters in the model names mean, and then return to what is happening here.

## Model Types

Each of the three letters in the model names have specific meanings and talk about different features of the groups (and how they relate to each other).

For this discussion we'll limit things to 2 groups to highlight how the different models and options influence the fitting.

### First Letter: Space

The first letter can be thought of as "how much space each group is allowed to take up." "E" means that each group should take up the same amount of "space" in the figure. "V" means that some groups might take up more space than others.

This is easier to see in a figure. Figure `r (fig=fig+1)` depicts what happens when change the first letter from E to V (ignore the other letters). We'll also assume there are only 2 groups.

```{r letter1, echo=T, fig.cap = "Comparing the Exx vs Vxx models", fig.height=4}

par(mfrow=c(1,2))
grps.mclst = Mclust(grps.LPA, model="EII", G=2, x = grps.BIC)
plot(grps.mclst, what="classification", ylim=rng, xlim=rng
     , colors=grpCols)
title(main="EII, 2")

grps.mclst = Mclust(grps.LPA, model="VII", G=2, x = grps.BIC)
plot(grps.mclst, what="classification", ylim=rng, xlim=rng
     , colors=grpCols)
title(main="VII, 2")
```

In the "EII" model (`r fig`), the circles describing the distributions are the exact same shape and size. In the VII model, the groups have different sizes. In particular, the "green group" has expanded, while the pink group has shrunk.

Note that size here does not refer to the number of observations in the group, but the amount of space the distribution takes up in the variable-space.

\newpage
### Second Letter: Relative amounts of each variable

In the two models in Figure `r fig`, the only thing that changes is the size of the circle. The general shape is the same (those models assume exactly equal variance in the X and Y directions).

The second letter controls whether or not the distributions can have different variances in the Reading and Math directions.

* **I**: The variance in the Reading and Math directions must be equal.
* **E**: The variance in the Reading and Math directions do not have to be equal, but the relative amounts of each variable must be the same across all groups.
* **V**: The variance in the Reading and Math directions do not have to be equal, and the proportions can be different for each group.

```{r letter2, echo=T, fig.cap = "Comparing the xIx vs xEx vs xVx models", fig.height=2.5}

par(mfrow=c(1,3))
grps.mclst = Mclust(grps.LPA, model="EII", G=2, x = grps.BIC)
plot(grps.mclst, what="classification", ylim=rng, xlim=rng
     , colors=grpCols)
title(main="EII, 2")

grps.mclst = Mclust(grps.LPA, model="EEI", G=2, x = grps.BIC)
plot(grps.mclst, what="classification", ylim=rng, xlim=rng
     , colors=grpCols)
title(main="EEI, 2")

grps.mclst = Mclust(grps.LPA, model="EVI", G=2, x = grps.BIC)
plot(grps.mclst, what="classification", ylim=rng, xlim=rng
     , colors=grpCols)
title(main="EVI, 2")
```

Figure `r (fig=fig+1)` illustrates these differences. You can see that the circles are perfect circles in the EII model on the left - this is because the "I" means that in every group, each group's distribution has equal variability in the two variables. In the middle panel, this constraint is relaxed and now there is a bit more variance in the Math direction than in the Reading direction - but the "E" forces the model to use the same ratio of variances for both groups. The right panel ("EVI") allows the variances to differ between the two groups. This produces relatively equal variances for the pink group, but more Math than Reading variance in the green group.

### Letter 3: Direction (covariance)

In the previous models, the groups assumed that Reading and Math were uncorrelated. What this means is that the circles and ovals are pointing up and down or left and right, rather than diagonally.

The third letter allows you to place constraints on the correlations between the variables. The coding scheme is similar to Letter 2.

* **I**: The covariance between Reading and Math must be 0. (Uncorrelated)
* **E**: The covariance between Reading and Math does not have to be 0, but it must be the same across all groups.
* **V**: The covariance between Reading and Math does not have to be 0, and the covariance can be different for each group.

```{r letter3, echo=T, fig.cap = "Comparing the xxI vs xxE vs xxV models", fig.height=2.5}

par(mfrow=c(1,3))
grps.mclst = Mclust(grps.LPA, model="EEI", G=2, x = grps.BIC)
plot(grps.mclst, what="classification", ylim=rng, xlim=rng
     , colors=grpCols)
title(main="EEI, 2")

grps.mclst = Mclust(grps.LPA, model="EEE", G=2, x = grps.BIC)
plot(grps.mclst, what="classification", ylim=rng, xlim=rng
     , colors=grpCols)
title(main="EEE, 2")

grps.mclst = Mclust(grps.LPA, model="EEV", G=2, x = grps.BIC)
plot(grps.mclst, what="classification", ylim=rng, xlim=rng
     , colors=grpCols)
title(main="EEV, 2")
```

Figure `r (fig=fig+1)` illustrates the difference in rotation. In the leftmost panel, all of the covariances are 0 and so the dotted lines form perfect pluses. In the middle panel, they are now rotated a little bit to reflect that there is a correlation between Reading and Math, but that correlation is assumed to be equal in both groups. It's fairly obvious that this does not work very well for the pink group. In the rightmost panel the correlations are allowed to vary between the groups, which leads to a positive correlation for the pink group and a negative one for the green group.

# OK, we have groups. Now what?

Once you have the results of the LPA, you probably want to examine the results and understand what they mean.

For this section we're going to stick with the "best fitting" VVV model with 3 groups.

The LPA results gives you several ways of examining the results.

## Characteristics of the "circles"

One thing we can do is just take a closer look at the characteristics of the 3 "bivariate distributions" that the LPA identified. Here are the three groups again (Figure `r (fig=fig+1)`):

```{r VVV3Plot, fig.cap="Depiction of the VVV,3 Results", fig.width=4, fig.heigth=4}
grps.mclst = Mclust(grps.LPA, model="VVV", G=3, x = grps.BIC)
plot(grps.mclst, what="classification", ylim=rng, xlim=rng
     , colors=grpCols)
title(main="VVV, 3")
```

The general summary associated with that can be obtained from the summary() function:

```{r distributionSummary, echo=TRUE}
summary(grps.mclst, parameters = T)
```

The first part of the output is about how well the data groups capture the variance in the data.

Next you get some information about the group sizes: group 1 has 5843 datapoints, group 2 has 9457, and group 3 has 6980. The mixing probabilities are just the proportions that are assigned to each group (so 5843 data points in group 1 represents ~26% of the data).

### Means

After the mixing probabilities we get the means for each distribution (I've rounded them here):

```{r distributionMeans, echo=TRUE}

# round the values of the means to make it a bit less busy
round(summary(grps.mclst, parameters = T)$mean, 2)

```

Each column represents one of the 3 distributions. So we can see that Group 1 has low reading and math scores (pink); Group 2 has a very high mean for Math, and somewhat average Reading scores (green group); Group 3 is average for both (blue).

### Covariances

Next we can look at the covariance structure (labelled Variances). (These are the variances and covariances, which underlie the correlations)

Each distribution has a matrix with the diagonals being the variances of Reading and Math, and the off-diagonal element being the covariance of Reading with Math.

```{r distributionCov, echo=TRUE}

round(grps.mclst$parameters$variance$sigma, 3)
```

Let's just consider the diagonal elements for the moment (variances). We can see that the pink group (G1 - the first matrix) has modest and relatively equal variances for both groups. Group 2 (green) has quite a lot more math variance than reading. Group 3 (blue) is generally high variance in both variables.

This is a good example of how "space" does not mean "group size". The blue group (Group 3) has far fewer observations than Group 2 (green), but it takes up a lot more space because the variances are considerably larger.

### Correlations
The covariances combine with the variances to give us the correlations. As a result, it's not obvious what the correlations are directly from the covariances matrices above. To get that we can use the cov2cor() function:

```{r distributionCors, echo=TRUE}

round(cov2cor(grps.mclst$parameters$variance$sigma[,,1]), 3) # Group 1
round(cov2cor(grps.mclst$parameters$variance$sigma[,,2]), 3) # Group 2
round(cov2cor(grps.mclst$parameters$variance$sigma[,,3]), 3) # Group 3
```

Correlations in the image show up as a combination of "stretchiness" in the oval, along with "tiltedness".

Here we see that G1 (pink) has a strong correlation (.66) - which fits the long, tilted shape of the distribution in the figure. G2 (green) has a modest, negative correlation (-.32), and G3 (blue) has a positive correlation that is a bit weaker than the pink group.

Of course, we could see all of that from the shape of the black ovals on the plot as well.

### Distribution of the fitted variables

Looking at how the variables that went into the data differ by groups can be useful to better understand what the LPA thinks represents different profiles.

### The sub-distributions

With that information, we can plot what each bivariate distribution looks like:

```{r groupDistributions, echo=TRUE, fig.cap="Multinomial distribution from the mixture models"}
xs = seq(-3,8,1/5)
dist1 = expand.grid(xs, xs) %>% 
  rename(Reading=Var1, Math=Var2) %>% 
  mutate(density=
           dmvnorm(.
                   
                   , mean=grps.mclst$parameters$mean[,1]
                   , sigma=grps.mclst$parameters$variance$sigma[,,1]
           )
  ) %>% 
  pivot_wider(id_cols=Reading, names_from=Math, values_from="density") %>% 
  select(-Reading)

dist2 = expand.grid(xs, xs) %>% 
  rename(Reading=Var1, Math=Var2) %>% 
  mutate(density=
           dmvnorm(.
                   
                   , mean=grps.mclst$parameters$mean[,2]
                   , sigma=grps.mclst$parameters$variance$sigma[,,2]
           )
  ) %>% 
  pivot_wider(id_cols=Reading, names_from=Math, values_from="density") %>% 
  select(-Reading)

dist3 = expand.grid(xs, xs) %>% 
  rename(Reading=Var1, Math=Var2) %>% 
  mutate(density=
           dmvnorm(.
                   
                   , mean=grps.mclst$parameters$mean[,3]
                   , sigma=grps.mclst$parameters$variance$sigma[,,3]
           )
  ) %>% 
  pivot_wider(id_cols=Reading, names_from=Math, values_from="density") %>% 
  select(-Reading)

par(mfrow=c(1,2))
# image2D(as.matrix(dist1), x = xs, y=xs, zlim=c(0,.35))
# image2D(as.matrix(dist2), x = xs, y=xs, zlim=c(0,.35))
# image2D(as.matrix(dist3), x = xs, y=xs, zlim=c(0,.35))
image2D(as.matrix(dist1+dist2+dist3), x = xs, y=xs, colkey=FALSE)
title("VVV 3 model distributions")
image2D(grps_z, x = as.numeric(rownames(grps_z)), y=as.numeric(colnames(grps_z))
        , ylim=range(xs), xlim=range(xs), colkey=FALSE)
title("Original data")


```

Figure `r (fig=fig+1)` clearly indicates that the model is doing quite well at reconstructing the distributions of our data.

## Characteristics on "non-LPA" variables.

However, we can also  ask questions about how members of each profile compare on variables that were not actually part of the LPA. There is a third variable in the dataset that we have not yet discussed - Need for Cognition. There are a few things we can do to look at how our profiles differ on other variables.

### Categorical approach

Looking at the last Figure `r fig-1`, it looks like each data point has been assigned to one of the three groups. We can use those assignments and  compare the three groups based on this classification.

The category assignments are stored in the grps.mclst object created by Mclust(...). We can link that to the original dataset.

```{r NeedForCogCategorical, echo=TRUE}

# The group assignments are stored in a "classification" vector in the
# grps.mclst object:
grps = data.frame(grps %>% select(Reading, Math, NeedForCog), Group = grps.mclst$classification)
rdat(grps, 5) # Custom function that randomly selects data
```

Using those "group" assignments, we can now compare the variables.

```{r NeedForCogCategoricalComparison, echo=TRUE}
grps %>% group_by(Group) %>% 
  summarize(N = n(), NeedForCog.mu = mean(NeedForCog), NeedForCog.sd = sd(NeedForCog))
```
In this case, Need for Cognition is highest in the pink group, and lowest in the blue group. The variances don't differ much.

We can even go a step further and conduct a regression analysis, or analysis of variance, or what have you:

```{r NeedForCogCategoricalRegression, echo=TRUE}
summary(lm(NeedForCog~Group, grps %>% mutate(Group=factor(Group))))
```

There you can see that both Group2 and Group 3 differ significantly from Group 1 (we have a large sample). pairwise.t.test is a quick way to compare multiple groups on a single variable with t-tests.

```{r NeedForCogCategoricalTTests, echo=TRUE}
pairwise.t.test(grps$NeedForCog, grps$Group)

```

All groups differ from each other.
\newpage

### Continuous approach

Most clustering techniques limit you to this approach - each datapoint is assigned to a category and you can use those categories to explore the data. One really nice feature of LPA is that it offers another more nuanced approach to exploring the data.

Let's revisit that figure again. If you look closely, there are a good number of points that fall sort of "between circles". Particularly between the blue and green groups, and between the blue and pink groups - there seem to be a lot of points that could probably be assigned to either distribution.

While the figures I've been producing above make very strong decisions about which Group each point belongs to, LPA is not quite that firm. In the figure below, instead of just the group assignment, we plot how "confident" the LPA analysis was in the group assignments. Technically we plot "uncertainty" - so large bubbles that the model is not very confident in its assignment for those points.

```{r VVV3PlotUncertainty, fig.cap = "Uncertainty of the VVV, 3 Model"}
plot(grps.mclst, what="uncertainty",ylim=rng, xlim=rng
     , colors=grpCols)
title(main="VVV, 3")
```

From Figure `r (fig = fig+1)`, you can see that because the distributions lie so close to each other, there is a fair amount of uncertainty about almost every point. At the more extreme points, there are some very small dots - these are observations with very low uncertainty.

The reason it can do this, is because LPA doesn't actually classify each data point, rather it takes those distributions that it calculated (means, and convariance matrices), and for every single datapoint, asks how likly it is to come from each of the distributions. From that it calculates the probability that each data point would fit into *each of the groups* separately. This means that for every observation, we have 3 probabilities, one for each of our groups.

Let's look at some of those data points:

```{r VVV3Probs}

data.frame(grps.mclst$z, Group=grps.mclst$classification) %>% slice(20826,1181,10058) %>% round(2)
```

Each column represents the probability that LPA would assign that particular data point to each group. The first entry is 99% likely to be group 2 (green), with little chance that it comes from the pink or blue distributions. The second entry is almost equally likely to be green or blue (2 or 3). The last is most likely to be pink, but could possibly be from blue.

What this means is that we don't have to treat each data point as if it belongs entirely to its assigned Group. We can think of the datapoint as being split amongst the groups in proportion to these probabilities.

- The first entry is clearly "green". There is little to no uncertainty.
- The second entry is highly uncertain, 57% green, 43% blue. (43% uncertainty)
- The third entry is 82% pink, 17% blue. (18% uncertain)

We can use these proportions to *weight* the values that we calculate in our other variables, rather than using the categorical variables.

This is a little bit tricky, but the basic approach looks like this:

```{r VVV3WeightedMeans, echo=T}
# The probabilities for each datapoint and each group are stored in $z of the 
# MClust object: grps.mclst$z. Each column refers to one group.

# To get the weighted mean of NeedForCog for Group 1, we use the first column of the
# probabilities.
NeedForCog.1=weighted.mean(grps$NeedForCog, w=grps.mclst$z[,1])
# Group 2:
NeedForCog.2=weighted.mean(grps$NeedForCog, w=grps.mclst$z[,2])
# Group 3:
NeedForCog.3=weighted.mean(grps$NeedForCog, w=grps.mclst$z[,3])

round(rbind(wtd = c(G1=NeedForCog.1, G2=NeedForCog.2, G3=NeedForCog.3),
      cat = c(G1=6.54, G2=4.98, G3=4.08)),2)


```

We can compare those to the "categorical" version and notice that they produce similar but not quite identical values.

It's also worth noting that by using weighted values rather than "all or nothing" categories, we change the "number" of datapoints in each Group.

```{r weightedNs, echo=T}

colSums(grps.mclst$z)
```


### A note on weighting

There are other calculations that can be weighted (e.g., variances and covariances, lots of statistical tests). I won't go over those here, but you can look for weighted versions of your analyses in R (many will require other packages). Weighting variables is not a trivial process, so be careful that you understand what is being weighted and how.

